---
title: "Motor Trend Car Road Tests on MPG"
author: "Gordon CHAN"
date: "2015/9/24"
output: pdf_document
---

# Introduction

In this report we performed data analyse on the *mtcars* dataset to explore the relationship between a sets of variables with the miles per gallon (MPG), with the aim to answer 2 questions:

1. Is an automatic or manual transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions

# Executive Summary

In this road test, 32 vehicles are tested on their Miles Per Gallon (MPG) and 11 variables are logged. From the data, a model (*mpg ~ wt + factor(cyl) + hp + am*) is fitted with MPG as the dependent variable, with 4 independent variables, weight, number of cylinders, horsepower and transmission identified.

# R-packages and Dataset

The *mtcars* dataset is loaded. For the analysis the *Caret* package is used.

```{r preparations}
# Load libraries
library(caret)
library(GGally)
library(gridExtra)
# Load dataset
data(mtcars)

```

# Exploratory Data Analysis

The *mtcars* dataset contained `r dim(mtcars)[2]` variables of `r dim(mtcars)[2]` models of cars tested.

```{r exploratory}
# Dataset dimensions
dim(mtcars)

```

From the documentation, the `r dim(mtcars)[2]` variables are explained:

Variable | Descripton
---------|-----------
mpg | Miles/(US) gallon
cyl | Number of cylinders
disp | Displacement (cu.in.)
hp | Gross horsepower
drat | Rear axle ratio
wt | Weight (lb/1000)
qsec | 1/4 mile time
vs | V/S
am | Transmission (0 = automatic, 1 = manual)
gear | Number of forward gears
carb | Number of carburetors

The pairs plot of the variables can be found in *Appendix 1*.

While **mpg** would be our dependent (outcome) variable, all others variables are potential independent variables. However, we have a special interst in **am** since it relates directly to the 2 question asked in the introduction.

# Regression modelling

Linear multivariable regression was performed. A rough model is fitted with **mpg** against all other variables.

```{r roughModel}
# Draft model1
modfit1 <- lm(mpg ~ factor(cyl) + disp + hp + drat + wt + qsec + 
                  vs + am + factor(gear) + carb, data = mtcars)
```

From the summary (*Appendix 2*), we can see that only a handful of variables showed a meaningfully large coefficient. Automatic variable selection was then performed to select the relavant variables. 

```{r selection}
# Skeleton model
modfit0 <- lm(mpg ~ 1, data = mtcars)
# Automatic variable selection by step function
stepfit <- step(modfit0, scope=list(lower=modfit0, upper=modfit1), direction="forward", trace = 0)
summary(stepfit)
```

Since we wish to explore the effect of transmission (am) on the mpg, the fitted model is compared with one that has a single variable *am*. From the ANOVA result, we could see that the difference is very siginificant, hence the null hypothesis that the variables *wt*, *cyl*, and *hp* are not contributing to the variability of the model is rejected.

```{r comparison}
# Baseline model
modfit.am <- lm(mpg ~ am, data = mtcars)
anova(modfit.am, stepfit)
```

If we look at the *Residual vs Fitted* plot of the model, we could see that the points are quite evenly and randomly scattered, while for the *Normal Q-Q* plot the points are lining quite neatly along the diagonal verifying that the residuals are randomly distributed. We could also observe some outliers in the higher end of mpg.

```{r diagnosis, fig.width = 5, fig.height = 5}
par(mfrow=c(2, 2))
plot(stepfit)
```

# Question 1: Is an automatic or manual transmission better for MPG?

In regard to the vehical transmission, the 1st question is relatively straight forward. A t-test is performed to compare the mean mpg of automatic / manual transmission group. We can see that there is a very significant difference between the groups, where cars with **manual transmission** has a higher MPG than those with automatic transmission. A boxplot can be found in Appendix 3.

```{r Q1, fig.width = 6, fig.height = 4}
t.test(mpg ~ am, data = mtcars)
```

# Question 2: Quantify the MPG difference between automatic and manual transmissions

Although from the t-test performed above, one might be tempted to quantify the difference between the groups by a simple subtraction. This would be incorrect as MPG is also affected by other factors as well. Here we should refer to the fitted model. From the *am* coefficient, we can conclude that cars with manual transmission get **1.80921** more MPG than an automatic.

# Conclusion

We have modelled the MPG of a given vehicle as *mpg ~ wt + factor(cyl) + hp + am*, in which it is affected by 4 variables:

1. Weight, a decrease in MPG of **2.49683** is expected per every increase of 1000lb in weight.
2. Number of cylinders, compared with those with 4 cylinders, a decrease in MPG of **3.03134** is expected for vehicle with 6 cylinders, and a decrease of **2.16368** for vehicle with 8 cylinders.
3. Horsepower, a decrease in MPG of **0.03211** is expected per every increase of 1 gross horsepower.
4. Transmission, an increase in MPG of **1.80921** is expected for manual vehicles.

# Appendicies

## Appendix 1: Pairs plot of variables in *mtcars*

```{r matrix, fig.width = 10, fig.height = 8}
# Pairs plot of mtcars
ggpairs(mtcars, 
        title = "Variables of mtcars dataset",
        size = list(corSize = 10, size = 2),
        alpha = 0.8,
        axisLabels = "internal")
```

## Appendix 2: Coefficient of fitted models

```{r modSummary}
summary(modfit1)$coef
```

## Apendix 3: Boxplot of MPG vs Transmissions

```{r Q1plot}
ggplot(data=mtcars, aes(am, mpg))+
    geom_boxplot(aes(fill=as.factor(am)))+
    labs(title ="Boxplot of Mean MPG vs Transmission", x = "Transmission", y = "MPG")+
    scale_fill_discrete(name = "Transmission", labels = c("Automatic", "Manual"))
```

